---
title: "Beer-Lambert Law"
output: ioslides_presentation
---
<style type="text/css">
body p, div, h1, h2, h3, h4, h5 {
color: black;
font-family: Modern Computer Roman;
}
slides > slide.title-slide hgroup h1 {
color: #8C1D40; <!--the maroon color-->
}
h2 {
color: #8C1D40; <!-- the maroon color-->
}
</style> <!-- end of defining font in various parts of slides -->

## Overview
This presentation demonstrates an application of statistics in chemistry, specifically the Beer-Lambert Law. In this presentation we will estimate the linear relationship between concentration and absorbance.

```{r setup, include=TRUE}
library(ggplot2)
library(dplyr)
library(plotly)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
## Simulated calibration data
Suppose we have a set of standard solutions with known concentrations (in mol/L) and their measured absorbances at a given wavelength.

```{r data-sim, echo=TRUE}
set.seed(42)
conc <- seq(0.1, 1.0, by=0.1)
absorb <- 0.25 + 1.2*conc + rnorm(length(conc), sd=0.05)
calib <- data.frame(conc, absorb)
head(calib)
```
## Beer-Lambert Law
The Beer-Lambert Law states:

$$
A = \varepsilon \, c \, l
$$
where:
- $A$ is absorbance
- $\varepsilon$ is molar absorptivity
- $c$ is concentration
- $l$ is path length (cm)

Calibration is often done using simple linear regression.

## Linear regression model
We can model absorbance as:

$$
A_i = \beta_0 + \beta_1 c_i + \varepsilon_i,
\qquad \varepsilon_i \sim N(0, \sigma^2)
$$

with estimates:

$$
\hat{\beta}_0, \hat{\beta}_1 = \arg\min \sum (A_i - (\beta_0+\beta_1 c_i))^2
$$

## Calibration curve
```{r gg-scatter, fig.width=4, fig.height=3}
p1 <- ggplot(calib, aes(x = conc, y = absorb)) +
geom_point(size = 2) +
geom_smooth(method = "lm", se = TRUE, color="blue") +
labs(title = "Calibration Curve", x = "Concentration (mol/L)", 
y = "Absorbance")
print(p1)
```

## Residuals histogram
```{r gg-resid, fig.width=4, fig.height=3}
fit <- lm(absorb ~ conc, data = calib)
resid_df <- data.frame(resid = residuals(fit))

p2 <- ggplot(resid_df, aes(x = resid)) +
geom_histogram(binwidth = 0.02, boundary = 0, closed = "left") +
labs(title = "Residuals of calibration model", x = "Residual", y = "Count")
print(p2)
```

## 3D interactive plot
Absorbance may depend on both concentration and wavelength. Below is a simulated 3D plot.

```{r plotly-3d, echo=FALSE}
wavelength <- rep(seq(400, 700, by=50), each=length(conc))
conc_long <- rep(conc, times=length(unique(wavelength)))
absorb3d <- 0.25 + 1.2*conc_long - 0.0005*(wavelength-550)^2 + rnorm(length(conc_long), sd=0.05)
plot_ly(x = ~conc_long, y = ~wavelength, z = ~absorb3d,
type = 'scatter3d', mode = 'markers', marker = list(size = 2)) %>%
layout(scene = list(xaxis = list(title = 'Concentration'),
yaxis = list(title = 'Wavelength (nm)'),
zaxis = list(title = 'Absorbance')))
```
