---
title: "Lab2"
author: "Kadee Books"
date: "9/14/2025"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(ggplot2movies)  
data(movies)
```

# Question 1
## Compute oldest and most recent movie years
```{r }
oldest <- min(movies$year, na.rm = TRUE)
newest <- max(movies$year, na.rm = TRUE)

cat("Oldest movie year:", oldest, "\n")
cat("Most recent movie year:", newest, "\n")
```

# Question 2
## Proportion of movies with and without a budget
```{r }
budget <- mean(!is.na(movies$budget))
no_budget <- mean(is.na(movies$budget))
budget
no_budget
```
## Top 5 most expensive movies
```{r }
movies %>%
  filter(!is.na(budget)) %>%
  arrange(desc(budget)) %>%
  select(title, year, budget) %>%
  head(5)
```
# Question 3
## Top 5 longest movies
```{r }
movies %>%
  arrange(desc(length)) %>%
  select(title, year, length) %>%
  head(5)
```
# Question 4
## Shortest and longest movie
```{r }
if ("Short" %in% names(movies)) {
  short_movies <- movies %>% filter(Short == 1)
  
  shortest_short <- short_movies %>% arrange(length) %>% slice(1)
  longest_short  <- short_movies %>% arrange(desc(length)) %>% slice(1)
  
  cat("Shortest short movie:", shortest_short$title, 
      "(", shortest_short$year, ") -", shortest_short$length, "minutes\n")
  cat("Longest short movie:", longest_short$title, 
      "(", longest_short$year, ") -", longest_short$length, "minutes\n")
}
```
# Question 5
## Movies of each genre in the database
```{r }
possible_genres <- c("Action","Animation","Comedy","Drama","Documentary","Romance","Short")
genres_present <- intersect(possible_genres, names(movies))

genre_counts <- movies %>%
  select(all_of(genres_present)) %>%
  summarise(across(everything(), \(x) sum(x, na.rm = TRUE)))

genre_counts_long <- tidyr::pivot_longer(genre_counts, everything(),
                                         names_to = "genre", values_to = "count")

ggplot(genre_counts_long, aes(x = genre, y = count, fill = genre)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Number of Movies by Genre", y = "Count")
```

# Question 6
## Average rating of all movies within each genre

```{r }
possible_genres <- c("Action","Animation","Comedy","Drama","Documentary","Romance","Short")
genres_present <- intersect(possible_genres, names(movies))

if (length(genres_present) > 0) {
  genre_ratings <- movies %>%
    select(rating, all_of(genres_present)) %>%
    tidyr::pivot_longer(cols = all_of(genres_present), 
                        names_to = "genre", values_to = "is_genre") %>%
    filter(is_genre == 1) %>%
    group_by(genre) %>%
    summarise(avg_rating = mean(rating, na.rm = TRUE))
  
  ggplot(genre_ratings, aes(x = genre, y = avg_rating, fill = genre)) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(title = "Average Rating by Genre", y = "Average Rating")
}
```

# Question 7
## Average rating of all movies within each genre between 2000-2005

```{r }
possible_genres <- c("Action","Animation","Comedy","Drama","Documentary","Romance","Short")
genres_present <- intersect(possible_genres, names(movies))

if (length(genres_present) > 0) {
  genre_ratings_2000_2005 <- movies %>%
    filter(year >= 2000 & year <= 2005) %>%
    select(rating, all_of(genres_present)) %>%
    tidyr::pivot_longer(cols = all_of(genres_present),
                        names_to = "genre", values_to = "is_genre") %>%
    filter(is_genre == 1) %>%
    group_by(genre) %>%
    summarise(avg_rating = mean(rating, na.rm = TRUE))

  ggplot(genre_ratings_2000_2005, aes(x = genre, y = avg_rating, fill = genre)) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(title = "Average Rating by Genre (2000–2005)", y = "Average Rating")
}
```

# Question 8
## Number of movies by genre per year (1990–latest year)

```{r }
possible_genres <- c("Action","Animation","Comedy","Drama","Documentary","Romance")
genres_present <- intersect(possible_genres, names(movies))

if (length(genres_present) > 0) {
  movies_1990 <- movies %>%
    filter(year >= 1990) %>%
    select(year, all_of(genres_present))

  genre_year <- movies_1990 %>%
    tidyr::pivot_longer(cols = all_of(genres_present),
                        names_to = "genre", values_to = "is_genre") %>%
    filter(is_genre == 1) %>%
    group_by(year, genre) %>%
    summarise(count = n(), .groups = "drop")

  ggplot(genre_year, aes(x = year, y = count, color = genre)) +
    geom_line(linewidth = 1) +
    theme_minimal() +
    labs(title = "Number of Movies by Genre (1990–Latest)", y = "Number of Movies")
}
```

# Question 9.1
## The year with highest average rating

```{r }
avg_rating_by_year <- movies %>%
  group_by(year) %>%
  summarise(avg_rating = mean(rating, na.rm = TRUE)) %>%
  arrange(desc(avg_rating)) %>%
  slice(1)

cat("Year with highest average rating:", avg_rating_by_year$year,
    "with rating", round(avg_rating_by_year$avg_rating, 2), "\n")
```

# Question 9.2
## Average movie length by year

```{r }
avg_length_by_year <- movies %>%
  group_by(year) %>%
  summarise(avg_length = mean(length, na.rm = TRUE))

ggplot(avg_length_by_year, aes(x = year, y = avg_length)) +
  geom_line(color = "blue", linewidth = 1) +
  theme_minimal() +
  labs(title = "Average Movie Length Over Time", y = "Average Length (minutes)")
```

# Question 9.3
## Distribution of IMDB ratings

```{r }
ggplot(movies, aes(x = rating)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of IMDB Ratings", x = "Rating", y = "Count")
```